---
title: "Practical Machine Learning Course Project"
author: "by Rebeccah Marsh"
output:
  html_document:
    fig_height: 9
    fig_width: 9
  pdf_document: default
---

## Background 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The goal of this project is to predict the manner in which the group performed the exercise and use the model to predict 20 different test cases.

## Preprocessing the data 
```{r, cache = T}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
```
### Download the data
```{r, cache = T}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```  
### Read the data
We read the downloaded data into two data frames:
```{r, cache = T}
trainRaw <- read.csv("./data/pml-training.csv")
testRaw <- read.csv("./data/pml-testing.csv")
dim(trainRaw)
dim(testRaw)
```
The training data set contains 19,622 observations and 160 variables. The testing data set contains 20 observations and 160 variables. 

Our objective is to presict the outcome "classe". 

### Clean the data
We clean the data, removing missing values and unnecessary variables:
```{r, cache = T}
sum(complete.cases(trainRaw))
```

Removing columns that contain NA missing values:
```{r, cache = T}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 
```  

Removing unnecessary columns:
```{r, cache = T}
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
```

Our cleaned training data set now contains 19,622 observations and 53 variables, and the testing data set contains 20 observations and 53 variables. 

### Slice the data
Next, we split the cleaned training set into a pure training data set (70%) and a validation data set (30%): 
```{r, cache = T}
set.seed(1234) # For reproducibile purpose
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```

## Data modeling
We fit a predictive model for activity recognition using the **Random Forest** algorithm (with **5-fold cross validation**) because it automatically selects important variables and is robust to correlated covariates & outliers in general:  
```{r, cache = T}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```

Next we test the performance of the model on the validation data set:  
```{r, cache = T}
predictRf <- predict(modelRf, testData)
confusionMatrix(testData$classe, predictRf)
accuracy <- postResample(predictRf, testData$classe)
accuracy
error <- 1 - as.numeric(confusionMatrix(testData$classe, predictRf)$overall[1])
error
```

Therefore, the estimated accuracy of the model is 99.42% and the estimated out-of-sample error is 0.58%.

## Prediction for the test data set
The final step is to apply the model to the testing data set provided:  
```{r, cache = T}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```  

## Visualizing the results
Correlation matrix: 
```{r, cache = T}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```

Decision tree:
```{r, cache = T}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) 
```